---
title: "HW6 Linear mixed-effect models"
output:
  html_document:
    df_print: paged
  pdf_document:
    latex_engine: xelatex
---
## 0. Packages and data
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(lme4)
library(lmerTest)
library(lingtypology) # only for linguistic mapping
library(broom)
```

### UPSID database
In this dataset we have number of consonants and vowels in 402 languages collected from UPSID database (http://www.lapsyd.ddl.ish-lyon.cnrs.fr/lapsyd/).  
* `language` - language  
* `area` - language area according to Glottolog (http://glottolog.org/)  
* `consonants` - the number of consonants in the language  
* `vowels` - the number of vowels in the language  

```{r, warning=FALSE, message=FALSE}
upsid <- read_csv("https://raw.githubusercontent.com/LingData2019/LingData2020/master/data/upsid.csv")
upsid
```

```{r}
# you can map the languages using the lingtypology package
# map.feature(upsid$language,
           # features = upsid$area,
           # label = upsid$language,
           # label.hide = TRUE)
```

In this work, you will fit a number of linear and linear mixed-effect models that predict the number of vowels in a language by the number of consonants and other variables.

## 1. Linear model

### 1.1 
Make a scatterplot of the number of consonants and the number of vowels using the `ggplot()`.   
```{r 1.1, echo = FALSE}
first_plot <- ggplot(data=upsid, aes(x=consonants, y=vowels)) + geom_point() + labs(x="consonants", y="vowels") + theme_bw()
first_plot
```

```{r}
# we use the theme_bw() theme, you can use other themes if you want
```
### 1.2
Fit the basic linear model `fit1` that predicts the number of vowels by the number of consonants. Look at the summary of the model.
```{r 1.2, include=FALSE}
# fit1 <- lm(...)
fit1 <- lm(data=upsid, vowels ~ consonants)
summary(fit1)
```

### 1.3
Is `consonants` a significant predictor? Write down YES or NO.
```{1.3}
YES
```

### 1.4
*no evaluation*  
To draw predictions on the graph, we may use `fortify` function. It adds column entitled `.fitted` (note period) that is calculated as a prediction of a model as well as some other columns.

Try this code and see the result:

```{r}
head(fortify(fit1))
```
Another option is to use `augment` function from `broom` library that works in a similar way.

```{r}
library(broom)
head(augment(fit1))
```

### 1.5
Use one of these function to add a line to the scatterplot 1.1. You have to use `geom_line` and pass the result of `augment` or `fortify` as argument `data` in `geom_line`. (E.g. `geom_line(data = <result of augment>, ...`))

```{r 1.5, echo=FALSE}
first_plot + geom_line(data = augment(fit1), aes(x = consonants, y = .fitted), color = "blue")
#
#  geom_line(data = augment(fit1), aes(x = consonants, y = .fitted), color = "blue")
```

## 2. Mixed-effect models
Let us look at the data with respect to the `area` groups. 

### 2.1
Re-build the scatterplot `1.1` coloring the points by `area`.
```{r 2.1, echo=FALSE}
second_plot <- ggplot(data=upsid, aes(x=consonants, y=vowels, color=area)) + geom_point() + labs(x="consonants", y="vowels") + theme_bw()
second_plot
# you can also draw stat_ellipse()
```

### 2.2
Use lmer() to fit the model `fit2` with random `area` group intercept. Your model is given by equation:

$$vowels = \beta_0 + \beta_1 \times consonants + u(area).$$
```{r 2.2, include=FALSE}
fit2 <- lmer(data=upsid, formula=vowels ~ consonants + (1|area))
summary(fit2)
```


### 2.3
Add the regression lines to the scatterplot `2.1` using `fortify` or `augment` methods.
```{r 2.3, echo=FALSE}
augment(fit2)
second_plot + geom_line(data = augment(fit2), aes(x = consonants, y = .fitted, color=area))
```

### 2.4
Interpret the results of the model. (Use `summary()` to get information about your model.) Is _consonansts_ variable still significant? How it changes after we added _area_ random effect? What can you say about lines on the graph? How are they located with respect to each other? Why?

```{2.4}
Yes, consonants are still significant.
The p-value became higher after we added area random effects.
The lines predict vowels using consonants for an ares, which gives us much more correct modelling.
They are parallel because of the random intercept - the formula is like 
  (b0 + random_constant_for_area) + b1 * X + e,
so the lines differ just for a constant.
```


## 3. Mixed-effect models with random slopes.
### 3.1
Fit the model `fit3` with random slope that depends on _area_. Your model is given by formula:

$$vowels = \beta_0 + (\beta_1 + u(area))\times consonants.$$
You can consult [this](http://bbolker.github.io/mixedmodels-misc/glmmFAQ.html#model-specification) manual to choose the correct syntax in your `lmer` formula.

```{r 3.1, include=FALSE}
fit3 <- lmer(data=upsid, formula=vowels ~ consonants + (0 + consonants|area))
summary(fit3)
```

### 3.2
Draw a figure with prediction lines:

```{r 3.2, echo=FALSE}
second_plot + geom_line(data = augment(fit3), aes(x = consonants, y = .fitted, color=area))
```

### 3.3
Interpret the results. Is _constants_ still significant? Why? What can you say about lines on the graph? How are they located with respect to each other? Why?

```{3.3}
Consonants are now not significant, p-value is at 1.
The lines should give us netter prediction of vowel count.
The lines now intersect in one spot because the formula now is like
  b0 + (b1 + random_constant_for_area) * X + e,
so the line angle now also is different for different areas.
```

## 4. Mixed-effect models with random intercept and random slopes.
Now let us assume we have both random intercept and random slope (and they are not correlated). Our model is of the form:

$$vowels = \beta_0 + u_1(area) + (\beta_1 + u_2(area))\times consonants.$$

### 4.1
Fit the model `fit4`.

```{r 4.1, include=FALSE}
fit4 <- lmer(data=upsid, formula=vowels ~ consonants + (0 + consonants|area) + (1|area))
summary(fit4)
```

### 4.2
Draw a figure with predictions:  

```{r 4.2, echo=FALSE}
second_plot + geom_line(data = augment(fit4), aes(x = consonants, y = .fitted, color=area))
```

### 4.3
Interpret the results. Is _constants_ significant? How can we interpret the difference between the prediction lines of these three models?

```{4.3}
Consonants are still not significant.
The difference of the lines exists because of the difference in formulas:
1) (b0 + random_constant_for_area) + b1 * X + e
2) b0 + (b1 + random_constant_for_area) * X + e,
3) (b0 + random_constant_for_area_1) + (b1 + random_constant_for_area_2) * X + e,
The first model tells us that there is smth like a base count of vowels for a language in an area.
The second model tells us that there is also a constant which can tell how the count of vowels grows based on the count of consonants.
The third model uses both this constants to show that they both make an effect on the count.
Thus we can say that if we take any area we can simply say that there will be at least (b0 + random_constant_for_area) vowels; we also can predict the growth of this count.
```
